---
title: "Dirichlet Process"
subtitle: "Patter Recognition Assignment"
author: 
      - "Name : Saheli Datta"
      
      - "Roll No : MD2213"
date: "2023-11-15"
toc: true
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage
# Introduction
The Dirichlet Process is a stochastic process used in Bayesian nonparametric modeling that acts as an infinite-dimensional generalization of the Dirichlet distribution. It's denoted as \( DP(\alpha, H) \), where \( \alpha \) is the concentration parameter and \( H \) represents a base distribution. Mathematically, it defines a distribution over probability measures, where for any finite partition of the space, the marginal probabilities follow a Dirichlet distribution, allowing for an unbounded number of clusters or components in clustering models. The distribution is defined as:

\[
G \sim DP(\alpha, H) \quad \text{and} \quad G(A_1, A_2, \dots, A_k) \sim \text{Dirichlet}\left(\alpha H(A_1), \alpha H(A_2), \dots, \alpha H(A_k)\right)
\]

Here, \( G \) is the random probability measure drawn from the Dirichlet Process, and \( A_1, A_2, \dots, A_k \) are disjoint measurable subsets of the space.

In Dirichlet Process modeling, each data point \( y_i \) is associated with a cluster parameter \( \theta_i \). The distinct values of cluster parameters \( \theta^*_i \) facilitate natural data grouping, making the Dirichlet process a potent method for cluster analysis. Particularly for multidimensional data, employing a mixture of multivariate normal distributions is prevalent to categorize observations into suitable clusters.

The full model can be expressed as:
$$
y_i \sim \mathcal{N}(y \, | \, \theta_i), \quad \theta_i = \{\mu_i, \Sigma_i\}, \quad \theta_i \sim G, \quad G \sim DP(\alpha, G_0)
$$

Where:
\begin{itemize}
  \item \( y_i \) represents the observation.
  \item \( \theta_i \) denotes the cluster parameter composed of mean (\( \mu_i \)) and covariance (\( \Sigma_i \)).
  \item \( \theta_i \) follows a distribution \( G \).
  \item \( G \) is drawn from a Dirichlet Process with concentration parameter \( \alpha \) and base distribution \( G_0 \).
\end{itemize}

# Objective
The objective is to determine the inherent number of clusters within the 'Faithful' dataset. This is achieved by iteratively performing cluster analysis using the Dirichlet Process model on the dataset while varying the number of iterations. The aim is to observe the number of clusters formed with each iteration. As the number of iterations increases substantially, the count of clusters is expected to converge towards a specific or stable value, indicating the probable number of natural clusters within the data.

# Data Description 
The 'Faithful' dataset is in the 'dirichletprocess' package. It is a data regarding Old Faithful Geysers in Yellowstone National Park, Wyoming, USA. There are total 272 observations on 2 variables.

- **eruptions:**  numeric	Eruption time in mins
- **waiting:** numeric	Waiting time to next eruption (in mins)

A closer look at faithful$eruptions reveals that these are heavily rounded times originally in seconds, where multiples of 5 are more frequent than expected under non-human measurement. 

\newpage
# Analysis of the Data
Here, in this section, we will look at how the data looks like through an exploratory analysis. Then we will move on to find out the true number of clusters that the data contains.

For that let us first import the necessary libraries and load the data:
```{r, message=FALSE, warning=FALSE}
# Installing and loading required packages
#install.packages("dirichletprocess")
library(dirichletprocess)
data("faithful")
set.seed(2023)
```
## Visualization and Summary of the data
First we plot the data. The scatterplot look like the following.
For that let us first import the necessary libraries and load the data:
```{r, message=FALSE, warning=FALSE}
library(ggplot2)
# Creating a scatter plot
ggplot(faithful, aes(x = eruptions, y = waiting)) +
  geom_point() +
  labs(x = "Eruptions", y = "Waiting", title = "Faithful Data")
```

From the above plot, we notice that, there are clearly 2 clusters in the data.

Given below the summary of the data:
```{r,message=FALSE,warning=FALSE}
summary(faithful)
```

## Cluster Analysis
We will consider 10, 50, 100, ..., 1000 number of iterations and will report the number of clusters formed in each iteration.

To do so, we first scale the data:
```{r,message=FALSE, warning=FALSE}
faithfulTrans <- scale(faithful)
```

For cluster analysis, we will use the 'DirichletProcessMvnormal()' function. It creates a Dirichlet mixture of multivariate normal distributions.

- **Case 1:** 10 Iterations
```{r, message=FALSE, warning=FALSE}
dp <- DirichletProcessMvnormal(faithfulTrans)
dp_10 <- Fit(dp, 10)
plot(dp_10)
```
Here we observe that only 1 cluster is formed.

- **Case 2:** 50 Iterations
```{r, message=FALSE, warning=FALSE}
dp_50 <- Fit(dp, 50)
plot(dp_50)
```
Here we observe that only 1 cluster is formed.

- **Case 2:** 100 Iterations
```{r, message=FALSE, warning=FALSE}
dp_100 <- Fit(dp, 100)
plot(dp_100)
```
Here we observe that only 3 cluster is formed.

- **Case 2:** 200 Iterations
```{r, message=FALSE, warning=FALSE}
dp_200 <- Fit(dp, 200)
plot(dp_200)
```
Here we observe that only 2 cluster is formed.

- **Case 2:** 1000 Iterations
```{r, message=FALSE, warning=FALSE}
dp_1000 <- Fit(dp, 200)
plot(dp_1000)
```
Here we observe that only 2 cluster is formed.

Below is the table for the number of clusters obtained by different number of iterations:

```{r,message=FALSE,warning=FALSE}
m=matrix(0,nrow=5,ncol=2)
m[,1]=c(10,50,100,200,1000)
m[1,2]=length(unique(dp_10$clusterLabels))
m[2,2]=length(unique(dp_50$clusterLabels))
m[3,2]=length(unique(dp_100$clusterLabels))
m[4,2]=length(unique(dp_200$clusterLabels))
m[5,2]=length(unique(dp_1000$clusterLabels))
colnames(m)=c("Iteration","Number of Clusters formed")
m
```